Investigating the alternative prey hypothesis with the POMP framework

### 2024-04-24 draft.Rmd
local searches

Partially observed Markov process models, also known as state-space models, hidden Markov models, and stochastic dynamical system models are probabilistic models used to describe the evolution of a system over time when some of the system's variables are "hidden" or unobservable. It consists of two main components: a latent *process model*, representing the unobservable states of the system, and a *measurement model*, linking the latent states to the observed data. By incorporating both observed and unobserved variables, POMP models allow researchers to make inferences about the underlying dynamics of the system. A simple visual representation of the process is depicted in the image below.

@@

```{r, echo=FALSE}
bird1 <- bird |> 
  dplyr::select(-R)

@@

A single execution of the `mif2` function applied to model provides a snapshot of the model's behavior under a specific set of initial conditions.

@@

Iterated Filtering (IF2) is implemented by a `foreach` loop which iterates 20 times, and in each iteration of the loop the IF2 function is applied to our model. This is a good way to explore the variability in outcomes due to initial conditions and other stochastic elements of the modeling process as well as to control the robustness. The variation in trajectories across the iterations suggests that the algorithm is exploring the parameter space and adjusting the estimates as it receives more information from the data.

@@

@@

>>>>

Partially observed Markov process models, also known as state-space models, hidden Markov models, and stochastic dynamical system models are probabilistic models used to describe the evolution of a system over time when some of the system's variables are "hidden" or unobservable [@ch10]. It consists of two main components: a latent *process model*, representing the unobservable states of the system, and a *measurement model*, linking the latent states to the observed data [@start]. By incorporating both observed and unobserved variables, POMP models allow researchers to make inferences about the underlying dynamics of the system. A simple visual representation of the process is depicted in the image below.

@@

```{r, echo=FALSE, warning=FALSE, message=FALSE}
bird1 <- bird |> 
  dplyr::select(-R)

@@

@@

A single execution of the `mif2` function applied to model provides a snapshot of the model's behavior under a specific set of initial conditions. Iterated Filtering (IF2) is implemented by a `foreach` loop which iterates 20 times, and in each iteration of the loop the IF2 function is applied to our model. This is a good way to explore the variability in outcomes due to initial conditions and other stochastic elements of the modeling process as well as to control the robustness. The variation in trajectories across the iterations suggests that the algorithm is exploring the parameter space and adjusting the estimates as it receives more information from the data.

@@

```{r, message=FALSE}
# bake(file="local_search.rds", {
#   foreach(i=1:20,.combine=c,
#     .options.future=list(seed=482947940)
#   ) %dofuture% {
#     mod |>
#       mif2(
#         Np=1000, Nmif=50,
#         cooling.fraction.50=0.5,
#         rw.sd = rw_sd(
#            alpha = 0.02, Beta = 0.02, gamma = 0.02,
#            a = 0.02, b = 0.02, c = 0.02,
#           sigma_obs = 0.02, logF_0 = ivp(0.1), logB_0 = ivp(0.1),
#          sigmaF = 0.02, sigmaB = 0.02, logRho = 0.02
#         ),
#         partrans = parameter_trans(
#              log = c('sigmaF', 'sigmaB', 'alpha', 'sigma_obs', 'a', 'gamma', 'Beta', 'b',
#               'c', 'logF_0', 'logB_0', 'logRho')),
#         paramnames=c("a", "b", "c", "gamma", "alpha", "Beta", "sigmaF", "sigmaB",
#                 "logF_0", "logB_0", "logRho", "sigma_obs"),
#         params = param_guess
#       )
#   }}) -> mifs_local

mifs_local |>
  traces() |>
  melt() |>
  ggplot(aes(x=iteration,y=value,group=.L1,color=factor(.L1)))+
  geom_line()+
  guides(color="none")+
  facet_wrap(~name,scales="free_y")
```

@@

```{r}
# bake(file="lik_local.rds",{
#   foreach(mf=mifs_local,.combine=rbind,
#     .options.future=list(seed=900242057)
#   ) %dofuture% {
#     evals <- replicate(10, logLik(pfilter(mf,Np=100)))
#     ll <- logmeanexp(evals,se=TRUE)
#     mf |> coef() |> bind_rows() |>
#       bind_cols(loglik=ll[1],loglik.se=ll[2])
#   } -> results
#   attr(results,"ncpu") <- nbrOfWorkers()
#   results
# }) -> results_local
# bind_rows(results_local) |> arrange(-loglik) |> write_csv("bird_params.csv")
# if (file.exists("CLUSTER.R")) {
#   source("CLUSTER.R")
# }
pairs(~loglik+a+alpha+b+Beta+c+gamma+logB_0+logF_0+logRho+sigmaB+sigmaF,data=results_local,pch=16)
```
